---
title: "Panel Data Problem Set 2"
author: "Ian Kennedy"
date: "`r format(Sys.Date(),'%m/%d/%Y')`"
output: 
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: no
    fig_caption: yes
    fig_width: 6
    fig_height: 4
header-includes: 
- \usepackage{graphicx}
- \usepackage{float}
- \usepackage{dsfont}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, autodep=TRUE,
                      cache=FALSE, cache.path="./cache/",
                      fig.align='center', fig.pos='H')
```

```{r, include=TRUE, message=FALSE}
library(knitr)
library(xtable)
library(tidyverse)
library(bookdown)
library(data.table)
```

Collaborators:  Devin Collins, Mark Igra, Ramin Jabbarli, and Hannah Lee

# Problem 1: Analyzing US House seat shares using ARMA

*[70 points total.]* Since 1963, the US House of Representatives has consisted 
of 435 elected voting members serving two-year terms.  Every seat in the House is up 
for election in November of even-numbered years to seat the Congress that will 
serve in the following two years. Thus, the 2016 election determined the 435 members of the House for the 115th Congress, serving from 2017–2018.  

We  will  study  the  evolution  of  the  time  series  of  the  number  of  
seats  won  by Democrats (or by independents who caucus with the Democratic Party) in each election held from 1963 to 2016 (a total of 28 observations). As substantive 
interest focuses on the party in control and the size of their majority, we will focus our analysis on these outcomes, where positive values indicate the size of a Democratic majority and negative values the size of a Republican majority.


Variable | Description
---------------- | ----------------------------------------
Congress | session of Congress (effectively a time index)
StartYear | the ﬁrst year of each two-year session
DemSenateSeats | the number of Democrats (and independents caucusing with the Democratic Party) elected to the Senate in this session of Congress
DemSenateMaj | the size of the Democratic Senate Majority,or DemSenateSeats minus 50
DemHouseSeats | the number of Democrats (and independents caucusing with the Democratic Party) elected to the House in this session of Congress
DemHouseMaj | the size of the Democratic Senate Majority, or DemHouseSeats minus 217
Midterm | whether this session was elected in a midterm election (1) or a presidential election (0)
DemPresident | whether the president on the last election day was a Democrat (1) or a Republican (0)
Unemployment | the monthly unemployment rate at the time of the election of this session of Congress
UnemDeviation | the difference between pre-election unemployment and mean unemployment, 1963–2016 (which was 6.075%)
Coattails | 1 if the presidency shifted to the Democrats on election day, 1 if the presidency shifted to the Republicans, and 0 if the party of the president was unchanged
PartisanMidterm | 1 for midterms in which the Democrats hold the presidency, 1 for midterms in which there is a Republican president, and 0 in presidential elections
PartisanUnem | equal to UnemDeviation when a Democrat is president, and to −1× UnemDeviation when a Republican is president  
Pre1994 | 1 if this Congress was elected before 1994, 0 otherwise

[Table 1. Codebook for Congressional Seats data. Data are in congress.csv, and 
are taken from the Bureau of Labor Statistics (unemployment) and Wikipedia (all other raw variables), or constructed from these data by your instructor.]  


We will also consider three possible inﬂuences on the size of the Democratic 
majority.  First, when one party sweeps the other party out of the presidency, they 
tend to bring in a wave of co-partisans to Congress who are “clinging to the 
president’s coattails.” Second, the party of the president tends to do badly in midterm elections (those that do not involve a presidential election; e.g., 2010, 2014, and 2018).  The usual explanation is that voters frustrated with the president cannot replace him, but can only offset his power with opposition in Congress.[^1]   Third, in all elections, voters tend to attribute economic performance to the party of the president.  For example, keeping unemployment below its long term average should help the president’s party at the expense of the opposition, and vice versa when unemployment is higher than usual.  

[^1]: More subtle theories exist; see Alberto Alesina and Howard Rosenthal, 1995, 
*Partisan Politics, Divided Government, and the Economy*, Cambridge University Press.  


These three explanations leave a lot out: for example, changes in the use of 
redistricting for partisan advantage, in the partisan composition of the electorate, and especially the transition of the Southern Democrats to the Republican Party. Because incumbency provides a strong advantage to sitting members of Congress, arguably many of these changes did not act gradually, but with a “bang” when a sudden shock caused many incumbents to lose or retire.  The shock in question is the 1994 midterm election; to account for the possibility it reﬂects a “structural break” in the level of the time series, we will also consider a control for whether our observations come before or after this watershed.  

In the ﬁle congress.csv, you will ﬁnd the variables described in Table 1. Examine the
data ﬁle, and note well the behavior of these variables over time. Then work 
through the following exercises:  

a.  *[10 points.]*  Plot the time series DemHouseMaj and plot its ACF and PACF. 
Perform augmented Dickey-Fuller and Phillips-Peron tests for unit roots. Describe
your ﬁndings, being sure to describe what kind(s) of time series process may be
at work. Now “demean” the data by period, removing the pre-1994 mean from
cases before 1994, and the post-1994 mean from cases after 1994. Make new time
series, ACF, and PACF plots. If 1994 represents a “structural break” in the 
level of the Democratic majority, what effect does that have on your diagnosis of the
behavior of the time series?  

b.  [10 points.] Fit an AR(0) regression to the time series `DemHouseMaj` controlling
for the covariates PartisanMidterm, PartisanUnem, and Coattails, which test the
three theories mentioned above.  Also control for Pre1994 to allow for a struc-
tural break. Present the results in a table, being sure to note the 
coefficients, their standard errors, the AIC for the entire model, the standard error of the regression, and the number of observations. Format the table nicely, as if for a paper, and describe what you have found substantively as well as you can.  

c.  [15 points.] Now ﬁt the following additional models and add them to the table
you made in part b: (i.)  an AR(1) model; (ii.)  an AR(2) model; (iii.)  an 
MA(1) model; (iv.) an ARMA(1,1) model. Make sure to include the same four controls
as   in part b.  Discuss the substantive and statistical similarities and differences
across all ﬁve ﬁtted models.

d.  [10 points.] Perform a rolling-windows cross-validation of all ﬁve models 
using a window of 20 periods and forecasting forward 3 periods.  Place in a table the
following goodness of ﬁt statistics for all ﬁve models: AIC, in-sample root mean
squared error, and the cross-validation mean absolute error (MAE) up to 1, 2, and 3 periods ahead, respectively, as well as the average of these threecross-validation
MAEs. Based on these statistics, select a ﬁnal “best” model.

e.  [25 points.] Using the model you selected in part d., forecast what will 
happen to the size of the Democratic majority in the US House in the 2018, 2020, and
2022 elections for three scenarios. For all three scenarios, assume the Democrats
recapture the presidency in 2020 and compute appropriate counterfactual values
of  PartisanMidterm and Coattails, and set Pre1994 to 0.  For unemployment,
assume the following:

Scenario | Counterfactual
-------- | --------
1 | unemployment stays at 4.6% for all three elections
2 | unemployment falls to 3.6% for all three elections
3 | unemployment rises to 5.6% for all three elections  

For each scenario, report or graph the predicted Democratic majority and its
95% conﬁdence (or predictive) interval for the 2018, 2020, and 2022 elections.  

Describe the substantive impact of your forecast results in as much detail as you
feel comfortable, as well as how much conﬁdence we should have in the forecasts.  

NB: As a check on your work, for each scenario and year also report the table of counterfactual covariate values you used to make your forecasts. Be very careful when constructing these values to capture to logic of the covariates; each one is tricky in its own way. To carry out the forecasts, you may use either predict() or the simcf library’s ldvsimev().  
 
Problem 2: Analyzing US Senate seat shares using ARMA  

*[30 points total.]* Since 1963, the US Senate has consisted of 100 elected 
voting members serving staggered six-year terms. Roughly one-third of the seats in the Senate are up for election in each even-numbered year. As a result, the Senate has three “classes” of seats. For example, the class of 2012 is up for re-election in 2018. Because 2012 was a good year for Democrats (due to Barack Obama’s coattails, among other factors), that means the Democrats have many seats to defend in 2018, putting them at a disadvantage relative to 2016.  

a.  [10 points.] Plot the time series DemSenateMaj and plot its ACF and PACF. 
Perform augmented Dickey-Fuller and Phillips-Peron tests for unit roots.  Now “demean” the data by period, removing the pre-1994 mean from cases before 1994, and the post-1994 mean from cases after 1994. Make new time series, ACF, and PACF plots and compare your results. Diagnosis the time series, accounting for the possibility of a structural break.  

b.  [10 points.]  Repurpose your code from Problem 1 to model the time series DemSenateMaj.  In particular, control for PartisanMidterm, PartisanUnem, Coattails, and Pre1994, and consider ﬁve models: an AR(0) model, an AR(1) model, an AR(2) model, an MA(1) model, and an ARMA(1,1) model. Recreate the two tables you made in Problem 1 (the table of coefficients and the table of goodness of ﬁt statistics) for the Senate data.  How do the substantive results compare to the House models?  

c.  [10 points.] Now estimate a sixth model: an AR(1)AR(1)₃. Add this model to the two tables you made in part b.  How well does the new model do?  What model is best overall?  Can you provide a substantive rationale for using either an MA(1) or an AR(1)AR(1)₃ model to model the US Senate, but not the House?  

